Insert the article title here

### 2021-06-24 fstats_pca.tex
figures

@@

\section{Results}
\begin{figure}[!ht]
	\includegraphics[width=\textwidth]{figures/fig_f3_data.pdf}
	\caption{\textbf{PCA and $F_3$-statistics} A: PCA of Western Eurasian data; the circle denotes the region for which $F_3(X; \text{Basque}, \text{Turkish})$ may be negative. Populations for which $F_3$ is negative are colored in red. B, E: $F_3$ approximated with two (blue) and ten (red) PCs versus the full spectrum. C,F: Contributions of PCs 1-10 to each $F_3$-statistic. D: PCA of World data set, color indicates value of $F_3(\text{Mbuti}; \text{Mozabite}, X)$. The black line shows the projection axis Mbuti-Mozabite, the gray lines indicates the projected position of each population. }
	\label{fig:f3}
\end{figure}

>>>>

\usepackage{url}

@@

\section{Results}
The theory outlined in the previous section suggests that $F$-statistics have a geometric interpretation on PCA plots. In this section, I use these interpretation in the analysis of human genetic variation data set. I use two data sets based on the ``Human Origins''-SNP set (597,573 SNPs). Both are subsets of the Reich lab compendium data set v44.3, downloaded from \url{https://reich.hms.harvard.edu/allen-ancient-dna-resource-aadr-downloadable-genotypes-present-day-and-ancient-dna-data}.

\paragraph{West-Eurasian data set}
This data set of 1,119 individuals from 62 populations contains  present-day individuals from the Eastern Mediterranean, Caucasus and Europe. It is frequently used as a basis of comparison for ancient genetic analyses of Western Eurasian individuals \cite{patterson2012}. Population genetic differentiation in this region is low and closely mirrors geography \cite{novembre2008}.

\paragraph{World Overview data set}
This data set of 638 individuals from 33 population contains individuals throughout the world, and is used as a sparse data set capturing much of global human genetic variation. This data set spans Africa, Eurasia and the Americas, and we might therefore expect the population structure to be much more sparse.

I perform analyses at the level of populations to ease presentation, and because it is an assumption of $F$-statistics that the genetic variation with sampled population is independent of the variation between samples that I am focusing on here. I use \texttt{admixtools 2.0.0} \url{https://github.com/uqrmaie1/admixtools} to compute a matrix of $F_2$-statistics between all populations. To obtain a PC-decomposition I use equation XXX and the \texttt{eigen} function in \texttt{R}, and compare them with the $F_3$ and $F_4$-statistics calculated using admixtools 2.

\paragraph{Admixture-$F_3$}
As a first step, I plot the first two principal components of the Westeurasian data set (Figure \ref{fig:f3}A). This PCA presents two parallel clines, one from the Levant and Arabia (``BedouinB'') to the Caucasus (``Abkhasian''), and a second one from Southern (``Sardinian'') to Northeastern Europe (``Mordovian''). In this context, I examine $F_3(X; Basque, Turkish)$, i.e. a statistic that aims to ask which populations can be represented as a mixture between a Southwestern (Basque) and Southeastern (Turkish) European population. The -- largely Southern European -- populations for which the point estimate of these $F_3$-statistic is negative are highlighted in red. They both fall close to the center of the $F_3$-circle, either defined on the first two (dark grey) or all PCs (light gray). However, many populations inside the circle on the first 2 PCs, including English, Sardinians and Canary Islanders have positive $F_3$-values, on higher PCs, showing that the first two PCs do not capture all the genetic variation related to population structure for this data set.

This is expected because for spatially continuous populations, PCA will not be sparse \cite{novembre2008a}.  Consequently, approximating $F_3$ by the first two or ten PCs (Figure \ref{fig:f3}B) only gives a coarse approximation of $F_3$, and from Figure \ref{fig:f3}C we see that many higher PCs contribute to $F_3$ statistics.

Thus, the main benefit of this PCA-plot is that it allows us to identify populations outside the circle (from the Levant and Caucasus), for which $F_3$ is guaranteed to be positive.

\paragraph{Outgroup-$F_3$}

\begin{figure}[!ht]
	\includegraphics[width=\textwidth]{figures/fig_f3_data.pdf}
	\caption{\textbf{PCA and $F_3$-statistics} A: PCA of Western Eurasian data; the circle denotes the region for which $F_3(X; \text{Basque}, \text{Turkish})$ may be negative. Populations for which $F_3$ is negative are colored in red. B, E: $F_3$ approximated with two (blue) and ten (red) PCs versus the full spectrum. C,F: Contributions of PCs 1-10 to each $F_3$-statistic. D: PCA of World data set, color indicates value of $F_3(\text{Mbuti}; \text{Mozabite}, X)$. The black line shows the projection axis Mbuti-Mozabite, the gray lines indicates the projected position of each population. }
	\label{fig:f3}
\end{figure}

### 2021-06-24 fstats_pca_lewontin.tex
figures

\subsubsection{$F_3$ and circles}
The $F_3$-statistic becomes more interesting; as outlines above we either think of $F_3$ as ``outgroup''-$F$-stats or as admixture $F$-stats. In the admixture case, we may ask the following question: given two source populations $X_1$, $X_2$, where would admixed populations on a PCA plot lie? From theory, we would expect it to lie between $X_1$ and $X_2$, with the exact location depending on sample sizes \cite{brisbin2012, mcvean2009}.

Formally, we would reject admixture if $F_3$ is negative, i.e. we are looking for the space
\begin{eqnarray}
2 F_3(X_x; X_1, X_2) &=& 2\langle  X_x - X_1, X_x - X_2 \rangle \nonumber\\
      &=& \normsq{X_x - X_1} + \normsq{X_x - X_2}  - \normsq{X_1 - X_2} \nonumber\\
      &<&0
\end{eqnarray}
By the Pythagorean theorem, $F_3 = 0 $ iff $X_1, X_2$ and $X_x$ form a right-angled triangle. Hence, the region where $F_3$ is zero is the circle with diameter through $X_1$ and $X_2$. If $X_x$ lies inside this circle, the angle is obtuse and $F_3$ is negative, otherwise it will be positive. Similarly, if we fix $X_1$ and $X_2$ and ask where on a 2D-PCA-plot $X_2$ would lie, this space is defined by all the points for which the angle between $X_1 X_x$ and $X_2 X_x$ is obtuse.

This highlights a potential identifiability issue with $F_3$: Since all values of $F_3$ that result in the same projection will give the same value; and multiple admixture events may result in the same $F_3$-value.

>>>>

\subsubsection{When is $F_3$ negative?}
The $F_3$-statistic becomes more interesting; as outlines above we either think of $F_3$ as ``outgroup''-$F$-stats or as admixture $F$-stats. In the admixture case, we may ask the following question: given two source populations $X_1$, $X_2$, where would admixed populations on a PCA plot lie? From theory, we would expect it to lie between $X_1$ and $X_2$, with the exact location depending on sample sizes \cite{brisbin2012, mcvean2009}.

Formally, we would reject admixture if $F_3$ is negative, i.e. we are looking for the space
\begin{eqnarray}
2 F_3(X_x; X_1, X_2) &=& 2\langle  X_x - X_1, X_x - X_2 \rangle \nonumber\\
      &=& \normsq{X_x - X_1} + \normsq{X_x - X_2}  - \normsq{X_1 - X_2} 
\end{eqnarray}
By the Pythagorean theorem, $F_3 = 0 $ iff $X_1, X_2$ and $X_x$ form a right-angled triangle. In a 2D-PCA plot, the region where $F_3$ is zero is the circle  with diameter $\overline{X_1X_2}$, and if $X_x$ lies inside this circle, $F_3(X_x; X_1, X_2) < 0$. If the

ball, the angle is obtuse and $F_3$ is negative, otherwise it will be positive. If we approximate the PCA-space in two dimensions, the $n$-ball corresponds to a circle.
